---
title: "R for Economics and Social Science Research"
author: "NE Milla, Jr."
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
subtitle: Key ANswer
---

```{=html}
<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 18px;
}
pre {
  font-size: 18px
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## WORKSHOP 4

1. Load the *rental* data set from the **wooldridge** package using the code below.
```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
library(tidyverse)
library(plm)
library(wooldridge)
data('rental')
head(rental)
rental %>% 
  select(city, year) %>% 
  table()

# 64 cities in 2 years
```




```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
rental %>%
  group_by(city) %>%
  summarize(lrent_mean = mean(lrent),
            var = sd(lrent)^2) %>% 
  arrange(-var)
```

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
rental %>%
  group_by(year) %>%
  summarize(lrent_mean = mean(lrent),
            var = sd(lrent)^2) %>% 
  arrange(-var)
```


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
rental %>%
  group_by(city) %>%
  summarize(lrent_mean = mean(lrent)) %>%
  left_join(rental) %>%
  ggplot(aes(x = reorder(as.character(city), city), 
             y = lrent)) +
  geom_point() +
  geom_line(aes(x = city, 
                y = lrent_mean), 
            col = "blue") +
  labs(x = "City", 
       y = "Rent") +
    theme(axis.text.x = element_text(angle = 90))
```


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
rental %>%
  group_by(year) %>%
  summarise(lrent_meanyr = mean(lrent)) %>%
  left_join(rental) %>%
  ggplot(aes(x = year, y = lrent)) +
  geom_point() +
  geom_line(aes(x = year, 
                y = lrent_meanyr), 
            col = "blue") +
  scale_x_continuous(labels = as.character(rental$year), 
                     breaks = rental$year) +
  labs(x = "Year", 
       y = "Rental") +
  theme(axis.text.x = element_text(angle = 90))
```



2. Fit an OLS model with *lrent* as response and *y90*, *lpop*, *lavginc*, and *pctstu* as predictors. You can use either the *lm()* or *plm()* function. Check the standard assumptions of a linear model. Which assumptions are violated, if any?

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
ols_mod <- lm(lrent ~ y90 + lpop + lavginc + pctstu,
              data = rental)

summary(ols_mod)
```

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
pooled_model_plm <- plm(lrent ~ y90 + lpop + lavginc + pctstu, 
                    data = rental, 
                    index = c("city", "year"), 
                    effect = "individual", 
                    model = "pooling")

summary(pooled_model_plm)
```


3. Using the same set of response and predictor variables fit a FE (within) model via LSDV approach
or using the *<tt>plm()* function. 

# Using LSDV approach


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
#City alone
lsdv_mod_city <- lm(lrent ~ y90 + lpop + lavginc + pctstu + factor(city),
              data = rental)

summary(lsdv_mod_city)
```



# Using plm()

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
fe_model_plm <- plm(lrent ~ y90 + lpop + lavginc + pctstu,
                    data = rental, 
                    index = c("city", "year"), 
                    effect = "individual", 
                    model = "within")

summary(fe_model_plm)
```



4. Use the *pFtest()* function to test for fixed effects. 

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
pFtest(fe_model_plm, pooled_model_plm)
```

Therefore, there are significant fixed effects. This implies that FE model is better than (pooled) OLS model.


5. Using the same set of response and predictor variables fit a RE model.


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
re_model_plm <- plm(lrent ~ y90 + lpop + lavginc + pctstu,
                    data = rental, 
                    index = c("city", "year"), 
                    effect = "individual", 
                    model = "random")

summary(re_model_plm)
```


6. Perform the Hausman test to determine whether a FE or RE model fits the data well.


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
phtest(fe_model_plm, re_model_plm)
```

Since the p-value < 0.05, we reject the null hypothesis (RE Model) and conclude that the FE model is more appropriate for this data.

7. Run the Breusch-Pagan Lagrange Multiplier test to determine if RE model fits better than (pooled) OLS model. 


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
plmtest(re_model_plm, 
        effect = "individual", 
        type = c("bp"))
```

Therefore, RE model is better than (pooled) OLS model.


In addition, FE model is also better than pooled OLS model as indicated by the small p-value in the output below.

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
plm::plmtest(fe_model_plm, 
        effect = "individual", 
        type = c("bp"))
```



8. Check all the assumptions of the final model. Re-estimate the coefficients, if necessary, based on the findings of the assumption checks.

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
# Test for heteroscedasticity
library(lmtest)
bptest(fe_model_plm)
```

Therefore, there is evidence of homoscedastic error variances.


```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
# Testing for autocorrelation
pbgtest(fe_model_plm)
```

Therefore, errors are not independent. Hence we apply robust standard errors as follows:

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
coeftest(fe_model_plm, 
         vcov = vcovHC(fe_model_plm,
                       type = "sss",
                       cluster = "group"))

#Uncorrected SE for comparison
data.frame("Uncorrected SE" = sqrt(diag(vcov(fe_model_plm))))
```


9. Interpret the coefficients of the best model.

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
coeftest(fe_model_plm, 
         vcov = vcovHC(fe_model_plm,
                       type = "sss",
                       cluster = "group"))

```

1. Assuming all else are equal, rent in 1990 is $e^{0.3855214} \approx 1.47$ times greater than the rent in 1980, or a 47% increase in rent from 1980 to 1990.

2. A 1% increase in population (pop) corresponds to about 0.07% increase in rent, holding all other predictors constant. This effect is too small and non-significant. Therefore increase in population does not significantly affect rent.

3. A 1% increase in average income (avginc) corresponds to approximately 0.31% increase in rent, assuming all else are equal.

4. A 1% increase in the percentage of student population (pctstu) is associated with approximately 0.01% increase in rent. [May not be practically significant!]  

