---
title: "WORKSHOP 2"
author: "NE Milla, Jr."
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
subtitle: Key Answer
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading packages

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
library(readxl)
library(tidyverse)
library(plotly)

```

# Loading the “World happiness data 2018-2023.xlsx”
```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
setwd("D:/Training and Seminars/2025/UP Visayas/Slides")
happiness <- read_excel("Data/World happiness data 2018-2023.xlsx")

happiness1 <- happiness %>% 
  rename("Year" = "year",
         "Country" = "country",
         "Region" = "region",
         "Happiness" = "happiness_score",
         "GDP" = "gdp_per_capita",
         "Social_Support" = "social_support",
         "Life_Expectancy" = "healthy_life_expectancy",
         "Freedom" = "freedom_to_make_life_choices",
         "Generosity" = "generosity",
         "Corruption" = "perceptions_of_corruption")
```

# Crosstab corruption level by region

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
happiness1 %>%
  filter(Year == 2023) %>%
  group_by(Region) %>% 
  summarize(M = round(mean(Corruption, na.rm=TRUE),2)) %>% 
  left_join(happiness1, by="Region") %>% #Using left_join() function
  select(Year, Region, Corruption, M) %>% 
  mutate("CorruptLevel" = if_else(Corruption <= M, "Below Average",
                                "Above Average"),
         Corruption=round(Corruption,2)) %>% 
  select(Region, CorruptLevel) %>% 
  group_by(Region) %>% 
  drop_na() %>% 
  count(CorruptLevel) %>% 
  as.data.frame() %>% 
  ggplot(aes(x = Region,
             y = n,
            fill = CorruptLevel)) +
  geom_bar(stat = "identity",
           position = "dodge",
           col = "black") +
  scale_fill_manual(values = c("Above Average" = "gray", 
                               "Below Average" = "lightblue")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) +
  scale_y_continuous(expand=c(0,0),
                     limits = c(0,150))+
  labs(x = "Region",
       y = "No. of countries",
       fill = "Corruption Level") +
  coord_flip() +
  theme_classic()
```

# Bar plot of the mean of healthy life expectancy for all three regions in Asia

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
happiness1 %>% 
  select(Region, Life_Expectancy) %>% 
  filter(Region %in% c("East Asia", "South Asia", "Southeast Asia")) %>% 
  group_by(Region) %>% 
  summarize(n = length(Life_Expectancy),
            Mean = mean(Life_Expectancy),
            SD = sd(Life_Expectancy)) %>% 
  mutate(SE = SD/sqrt(n)) %>% 
  as.data.frame() %>% 
  ggplot(aes(x=Region, y = Mean)) +
  geom_col(fill= "lightblue", 
           col="black", 
           width = 0.5) +
  geom_errorbar(aes(ymin = Mean-SE, 
                    ymax = Mean+SE),
                size = .7,
                width = 0.15) + 
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,1))+
  theme_classic()
```

# scatter plots of GDP and happiness score: 2022 & 2023

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
g1 <- happiness1 %>% 
  select(Year, GDP, Happiness) %>% 
  filter(Year %in% c(2022, 2023)) %>% 
  mutate(Year = factor(Year)) %>% #To convert year as categorical variable
  ggplot(aes(x = GDP,
             y = Happiness,
             color = Year)) +
  geom_point(size = 1.3,
             alpha = 1.5) +
  labs(x = "GDP",
       y = "Happiness") +
  theme(legend.position = "none")+ #To remove the legend
  facet_wrap(~factor(Year)) +
  theme_classic() 
  
ggplotly(g1)
  
```

# Animated plot of freedom to make life choices and happiness score

```{r eval=FALSE}
g2 <- happiness1 %>% 
  select(Year, Freedom, Happiness) %>% 
  ggplot(aes(x = Freedom, 
             y = Happiness)) +
  geom_point(aes(frame = as.integer(Year))) +
  theme_classic()

ggplotly(g2)
```


# Animated plots using gganimate

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
library(gganimate)
happiness1 %>% 
  select(Year, Freedom, Happiness) %>% 
  ggplot(aes(x = Freedom, 
             y = Happiness)) +
  geom_point(color = "maroon")  +
  labs(title = 'Year: {frame_time}') +
  transition_time(as.integer(Year)) + #To ensure the years will be treated as integer
  theme_classic()
```


# Density plot of happiness index

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
happy <- happiness1 %>% 
  ggplot(aes(x = Happiness)) +
  geom_density(fill = "lightblue") +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic()
```

# Density plot of GDP

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
gdp <- happiness1 %>% 
  ggplot(aes(x = GDP)) +
  geom_density(fill = "lightblue") +
  scale_y_continuous(expand = c(0,0))+
  theme_classic()
```


# Density plot of Freedom

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
free <- happiness1 %>% 
  ggplot(aes(x = Freedom)) +
  geom_density(fill = "lightblue") +
  scale_y_continuous(expand = c(0,0))+
  theme_classic()
```


# Density plot of Life Expectancy

```{r}
life <- happiness1 %>% 
  ggplot(aes(x = Life_Expectancy)) +
  geom_density(fill = "lightblue") +
  scale_y_continuous(expand = c(0,0))+
  theme_classic()
```


# Combining the four plots

```{r message=FALSE, warning=FALSE, class.output="bg-warning", class.source="bg-info"}
library(patchwork)
(happy + gdp) / (free + life)
```

